Functional timeseries reveal no systemic shifts for vertebrate communities

### 2021-06-18 paper/manuscript.Rmd
Add header
Update figures, revamp discussion

journal: "Ecography"
date: "`r Sys.Date()`"
bibliography: refs.bib
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: #word_document
  rticles::elsevier_article:
  includes:
   in_header: preamble.tex
editor_options: 
  markdown: 
    wrap: 72
---

@@

Functional diversity offers a potentially powerful alternative for
detecting and describing community change by providing a mechanistic
link between the organisms in a system and the processes they perform
[@mcgill2006]. By describing the functional trait space rather than
species, functional diversity metrics capture the disproportionate
impact of losses or gains of functionally unique species. Functional
diversity metrics are therefore particularly well-suited for addressing
change in communities with constant species richness.

Loss of functional diversity is also linked to impacts on critical
ecosystem and community processes. For example, there is increasing
evidence that functional diversity is a better predictor of ecosystem
function than species-based metrics [@cadotte2011; @gagic2015], and that
different facets of functional diversity play essential roles in
maintaining ecosystem stability [@craven2018; @morin2014]. Further,
maintenance of functional diversity is key for maximizing adaptive
potential of communities under climate change (citation).

<!-- Functional loss is often implicit in work linking species loss to impacts on ecosystem processes despite a wide range of documented relationships between functional and species change [@biswas2011]. Examining functional and species change side by side across a variety of communities -->

<!-- What can we say about how global change is impacting:stability, resilience, function, assembly, ecosystem processes (and ecosystem services) - need a functional grounding for all those concepts. Link specific patterns in FD metrics to impacts on those concepts. -->

<!-- paragraph about what we already know about global functional changes -->

Assessments of broad-scale temporal change in functional diversity have
previously been limited by a lack of functional trait data. The majority
of work has therefore focused largely on system-specific studies with
traits collected *in situ*. Ongoing efforts to assemble functional
traits for a variety of taxa have made synthesis of existing community
assemblage data and functional traits possible for the first time,
providing initial insights into the ways functional diversity changes on
a broad scale for specific taxa (e.g. fish[@trindade-santos2020], birds
[@jarzyna2016; @barnagaud2017]).
<!-- so what do these studies say? does that go here? --> However, to
date there has been no cross-taxa assessment of temporal functional
change for a broad geographic and taxonomic extent.

@@

We obtained mammal, bird, and amphibian timeseries from the Biotime
database, a global repository of high quality community assemblage
timeseries. All studies included in the database follow consistent
sampling protocols and represent full assemblages rather than population
of single species [@dornelas2018]. Following best practices for the
database [@blowes2019], studies with multiple sample locations were
split into individual timeseries following a standardized spatial scale.
Scale was set by a global grid with cell size determined based on the
sample extent of studies with only a single location (see
[@dornelas2018] for details on how sample extents were defined), with
the area of each cell set to one standard deviation away from the mean
of the single extent locations. All samples from a study within a single
cell were considered to be a single timeseries, and species abundances
were combined for all samples.

We gathered trait data from the Elton Trait Database (mammals and birds;
[@wilman2014]) and Amphibio (amphibians, [@oliveira2017]). These
databases include species-level means for traits that best define
species' function in the community including body size, diet, and
behavioral characteristics<!-- what's in the amphibian database? -->.
For the full list of traits included in the analysis see the
supplemental materials.
<!-- limitations of species level trait data? -->

In order to ensure taxonomic consistency across datasets, Biotime
species were paired with trait data based on their species identifier
from the Integrated Taxonomic Information System database (citation),
obtained through the taxadb R package [@norman2020]. If more than one
species in the assemblage data resolved to the same identifier,
observations were considered the same species. For trait data, traits
for all species of the same identifier were averaged. Only studies with
at least 75% trait coverage were included. In order to have a sufficient
number of species to calculate functional diversity metrics, years with
fewer than 4 species observed were excluded.

@@

Our final dataset included 2,439 timeseries from 54 studies in 21
countries and 15 biomes <!-- list traits -->. Only four studies
(consisting of 11 timeseries) came from Amphibian studies due to the
limited availability of amphibian timeseries and low species richness
values for assemblages. For a full breakdown of studies and their
characteristics, see the supplement.

@@

Functional diversity metrics were calculated using the *dbFD* function
from the *FD* R package. Here we report functional richness (*FRic*),
functional evenness (*FEve*), and functional diveregence (*FDiv*) which
together describe three complementary characteristics of the functional
space [@mason2005]. *FRic* assesses the volume of the trait space
occupied by species in the community. *FEve* describes how species are
distributed across the trait space and how abundance is distributed
across species. *FDiv* measures the degree to which species and their
abundances maximize differences in the functional space. Higher values
of *FDiv* therefore correspond to communities where many highly abundant
species are on the edges of the trait space.

@@

<!-- why do some studies not have blup estimates for FRic?-->

@@

We found no significant temporal trend in species richness or functional
diversity metrics for either observed or standardized metrics. There was
also no trend for studies broken down by taxa, biome or realm. We did
find a significant overall decrease in Jaccard similarity, indicating
increasing turnover through time.

At the study level, the majority of studies (78%) experienced a
significant temporal change in species richness and all functional
diversity metrics. 92% of sites experienced a significant trend for
species richness, 88% for functional functional richness, 96% for
functional divergence, 88% for functional evenness, and 35% for Jaccard
similarity. Only 4 (cite studies) out of 54 studies experienced no
significant trend in species richness and were also characterized by
significantly decreasing functional evenness. There was only one
significant pairwise correlation between the metrics, a positive
relationship between Functional Divergence and Functional Evenness,
indicating that communities experiencing an increase in Divergence
frequently experience an increase in Evenness.

@@

Our study represents the largest broad-scale multi-taxa assessment of
functional change through time, giving a first look at overall trends in
functional diversity and the range of changes happening in mammal, bird,
and amphibian communities. Consistent with previous species-based
syntheses, we found no overall trend in species diversity accompanied by
increasing turnover through time [@dornelas2018]. While nearly all of
the studies included in our data experienced a significant increase or
decrease in species richness, trends were observed in equal numbers
resulting in an overall trend not significantly different from zero.

Surprisingly, we did not detect an overall trend in any of the observed
or corrected functional diversity metrics. Similar to species richness,
communities are experiencing a wide range of functional shifts that
balance each other out. While our dataset likely under-samples areas
with most direct human impact, anthropogenic change is hypothesized to
impact even the most remote systems. Conservatively, our findings
therefore indicate that in unperturbed mammal, bird, and amphibian
communities there is no evidence of systemic functional richness loss or
functional change. A lack of trend for data broken down by realm, biome,
and taxonomic group gives further evidence that directional functional
change is absent from all systems observed in our dataset.

Individual studies in our dataset demonstrate the variety of species and
functional shifts that can happen concurrently in a community. With only
one significant pairwise relationship between Functional Divergence and
Functional Evenness, all other possible combinations of species and
functional change were observed in at least one study. Notably, there is
no relationship between species richness change and change in functional
richness, meaning that even for studies that may be experiencing species
losses there is not necessarily an accompanying loss of functional
richness and in some cases there may even be an increase. These
communities are gaining new species that are functionally distinct
enough to make up for functional losses due to species losses. This
pattern would be further magnified if the species lost are not
functionally distinct members of the assemblage.

@@

Our results have important implications for how we approach work linking
species loss to ecosystem processes, which often implicitly assume
concurrent functional loss. There is no evidence of directional
functional loss or even functional shifts accompany species loss in
these natural communities.
<!-- clarify that this is about relationship between concurrent patterns in a single community -->
Building on a rich body of work linking species diversity changes to
changes in ecosystem function, resilience, and stability (citations), it
will be essential to tease apart effects due to species change from
those due to functional shifts if we want to infer impacts in natural
communities.

@@

>>>>

journal: "Ecology Letters"
date: "`r Sys.Date()`"
bibliography: refs.bib
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: #word_document
  rticles::elsevier_article:
  includes:
   in_header: preamble.tex
editor_options: 
  markdown: 
    wrap: sentence
---

```{r message=FALSE, include = FALSE}
library(kableExtra)
library(magrittr)
library(tidyverse)
#library(printr)
library(rticles)
library(knitr)

knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)

# no sci notation integers pleases
options(scipen=999)
```

@@

Functional diversity offers a potentially powerful alternative for detecting and describing community change by providing a mechanistic link between the organisms in a system and the processes they perform [@mcgill2006].
By describing the functional trait space rather than species, functional diversity metrics capture the disproportionate impact of losses or gains of functionally unique species.
Functional diversity metrics are therefore particularly well-suited for assessing the implications of community shifts in the full range of species change scenarios.

Trends in functional composition also have important implications for how we assesses potential changes in ecosystem stability, function, and resilience.
Functional loss is often implicit in work linking biodiversity loss to impacts on ecosystem processes despite a wide range of documented relationships between functional and species change in natural communities [@biswas2011].
<!-- For example, hypotheses underpining the biodiversity ecosystem function relationship: insert function/stability theory here. --> Where functional diversity is explicitly considered, there is increasing evidence it is a better predictor of ecosystem function than species-based metrics [@cadotte2011; @gagic2015], and that different facets of functional diversity play essential roles in maintaining ecosystem stability [@craven2018; @morin2014].
Tracking the relationship between species and functional changes across a wide variety of communities therefore provides an opportunity to assess not only the prevalence of concurrent species and functional loss, but also the evidence underpinning hypothesis between biodiversity loss and ecosystem processes.

<!-- paragraph on maintenence of functional structure through time -->

Assessments of broad-scale temporal change in functional diversity have previously been limited by a lack of functional trait data.
The majority of work has therefore focused largely on system-specific studies with traits collected *in situ*.
Ongoing efforts to assemble functional traits for a variety of taxa have made synthesis of existing community assemblage data and functional traits possible for the first time, providing initial insights into the ways functional diversity changes on a broad scale for specific taxa [e.g. fish @trindade-santos2020, birds @jarzyna2016; @barnagaud2017].
<!-- so what do these studies say? does that go here? --> However, to date there has been no cross-taxa assessment of temporal functional change for a broad geographic and taxonomic extent.

@@

We obtained mammal, bird, and amphibian timeseries from the Biotime database, a global repository of high quality community assemblage timeseries.
All studies included in the database follow consistent sampling protocols and represent full assemblages rather than populations of single species [@dornelas2018].
Following best practices for the database [@blowes2019], studies with multiple sample locations were split into individual timeseries following a standardized spatial scale.
Scale was set by a global grid with cell size determined based on the sample extent of studies with only a single location [see @dornelas2018] for details on how sample extents were defined), with the area of each cell set to one standard deviation away from the mean of the single extent locations.
All samples from a study within a single cell were considered to be a single timeseries, and species abundances were combined for all samples.

```{r echo = FALSE, results = "asis"}
tibble(Taxa = c("Mammals", "Birds", "Amphibians"),
       Timeseries_Count = c(48, 2380, 11),
       Species_Count = c(184, 700, 184),
       Trait_Source = c("Elton Traits", "Elton Traits", "Amphibio"),
       Traits = c("body mass, diet, active diel period", "body mass, diet, nocturnality, \nforest foraging strata, pelagic specialist", "habitat, diet, active diel period, \nactivity seasonality, body mass, body length, \nmin maturation size, max maturation size, \nmin offspring size, max offspring size, \nreproductive output, breeding strategy")
) %>%
  mutate_all(linebreak, align = "l") %>% 
  knitr::kable(format = "latex", col.names = c("Taxa", "Number of \nTimeseries", "Number of \nSpecies", "Trait Source", "Traits"), escape = FALSE, 
               caption = "Summary of the data in the final trait database.") %>%
   kable_styling(latex_options= c("scale_down", "hold_position")) #%>%
   #kableExtra::column_spec(2, width = "5cm") #%>%
  # kableExtra::column_spec(4, width = "10cm")
```

We gathered trait data from the Elton Trait Database [mammals and birds, @wilman2014] and Amphibio [amphibians, @oliveira2017].
These databases include species-level means for traits that best define species' function in the community including body size, diet, and behavioral characteristics.
For the full list of traits included in the analysis for each taxa see Table 1.
<!-- limitations of species level trait data? -->

In order to ensure taxonomic consistency across datasets, Biotime species were paired with trait data based on their species identifier from the Integrated Taxonomic Information System database (citation), obtained through the taxadb R package [@norman2020].
If more than one species in the assemblage data resolved to the same identifier, observations were considered the same species.
For trait data, traits for all species of the same identifier were averaged.
Only studies with at least 75% trait coverage were included and observations for species with no trait data were excluded.
In order to have a sufficient number of species to calculate functional diversity metrics, years with fewer than 5 species observed were also excluded.

@@

Our final dataset included 2,443 timeseries from 53 studies in 21 countries and 15 biomes and 13 different traits (Fig 1).
Only four studies (consisting of 11 timeseries) came from Amphibian studies due to the limited availability of amphibian timeseries and low species richness values for assemblages (Table 1).
For a full breakdown of studies and their characteristics, see the supplement.

```{r taxa-map, fig.cap="Map of timeseries locations with points color corresponding to taxa.", echo=FALSE}
knitr::include_graphics(here::here("figures/study_map.jpeg"))
```

@@

Functional diversity metrics were calculated using the *dbFD* function from the *FD* R package.
Here we report functional richness (*FRic*), functional evenness (*FEve*), and functional diveregence (*FDiv*) which together describe three complementary characteristics of the functional space [@mason2005].
*FRic* assesses the volume of the trait space occupied by species in the community, with higher values indicating communities with species of more extreme trait values.
*FEve* describes how species are distributed across the trait space and how abundance is distributed across species.
Higher values of *FEve* indicate more even spacing of species in the trait space and individuals across species.
*FDiv* measures the degree to which species and their abundances maximize differences in the functional space.
Higher values of *FDiv* therefore correspond to communities where many highly abundant species are on the edges of the trait space.
<!-- use these metrics because they're common, not because of their results -->

@@

```{r timeseries-plot, fig.cap="Plots of timeseries-level trends with line color corresponding to climatic region, with data points in grey and the overall metric mean in black for A) log species richness, B) Jaccard similarity, C) Functional Richness SES, D) Functional Divergence SES, and E) Functional Evenness SES", echo=FALSE}
knitr::include_graphics(here::here("figures/3met_long.jpeg"))
```

@@

<!-- Despite significant trends at the study level for the majority of studies,-->

We found no significant overall trend in species richness or functional diversity metrics (Fig 2).
There was also no temporal trend for studies broken down by taxa, biome or realm.
We did find a significant overall decrease in Jaccard similarity, indicating increasing turnover through time.
Non-significant overall trends indicate that an equal number of studies experienced increasing and decreasing trends in those metrics.

At the study level, the majority of studies (86%) experienced a significant temporal change in species richness and all functional diversity metrics.
92% of studies experienced a significant trend for species richness, 88% for functional richness, 98% for functional divergence, 96% for functional evenness, and 30% for Jaccard similarity.
Only 4 (cite studies) out of 53 studies experienced no significant trend in species richness and were also characterized by significantly decreasing functional evenness.
There was only one significant pairwise correlation between the metrics, a positive relationship between Functional Divergence and Functional Evenness, indicating that communities experiencing an increase in Divergence frequently experience an increase in Evenness.

@@

<!-- Our study represents the largest broad-scale multi-taxa assessment of functional change through time, giving a first look at overall trends in functional diversity and the range of changes happening in mammal, bird, and amphibian communities. Consistent with previous species-based syntheses, we found no overall trend in species diversity accompanied by increasing turnover through time [@dornelas2018]. While nearly all of the studies included in our data experienced a significant increase or decrease in species richness, trends were observed in equal numbers resulting in an overall trend not significantly different from zero. -->

<!-- Surprisingly, we did not detect an overall trend in any of the functional diversity metrics. Similar to species richness, communities are experiencing a wide range of functional shifts that balance each other out. While our dataset likely under-samples areas with most direct human impact, anthropogenic change is hypothesized to impact even the most remote systems. Conservatively, our findings therefore indicate that in unperturbed mammal, bird, and amphibian communities there is no evidence of systemic functional richness loss or functional change. A lack of trend for data broken down by realm, biome, and taxonomic group gives further evidence that directional functional change is absent from all systems observed in our dataset. -->

Our study represents the largest broad-scale multi-taxa assessment of functional change through time to date, giving a first look at overall trends in functional diversity and the range of changes happening in mammal, bird, and amphibian communities.
Surprisingly, we did not detect an overall trend in any of the calculated functional diversity metrics.
While nearly all of the studies included in our data experienced a significant increase or decrease in at least one of the metrics, trends were observed in equal numbers resulting in an overall trend that was not significantly different from zero.
As with previous species-based syntheses, we also found no overall trend in species diversity accompanied by increasing turnover through time [@dornelas2018], indicating that non-significant trends in functional metrics may be consistent with similar well-documented species derived trends.
We find no evidence of systemic functional richness loss or functional change.
A lack of trend for data broken down by realm, biome, and taxonomic group gives further evidence that directional functional change is absent from all systems observed in our dataset.

Individual studies in our dataset demonstrate the variety of species and functional shifts that can happen concurrently in a community.
With only one significant pairwise relationship between study-level functional divergence slopes and functional evenness slopes, all other possible combinations of species and functional change were observed in at least one study.
Notably, there is no relationship between species richness change and change in functional richness, meaning that communities experiencing species losses do not necessarily experience accompanying functional richness, and in some cases may even experience an increase.
<!-- need to put a disclaimer about what "increasing" means here - it's increased richness relative to the expectation for a community of that size --> These communities gain species that expand the boundaries of the trait space while losing interior species.

@@

*Methodological Limitations*

This study represents best practices in biodiversity synthesis and functional diversity analysis, however methodological limitations may still be obscuring potential systematic shifts in functional composition.
First, as previously acknowledged, the Biotime database is not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias.
While both of these challenges are pervasive in Ecology generally and could have potentially insidious implications for synthesis work, we can only use the data we have.
We do however emphasize that our study included many timeseries that experienced a significant species loss (\~ 52%), indicating that this is not simply a synthesis of pristine systems.

Second, despite using the most comprehensive trait databases for these taxa, we were still limited to species-level means of the traits deemed important by database creators.
The importance of intraspecific variation is well documented [@desroches2018; @violle2012], however individual-level traits are rarely collected along side monitoring data, especially for the longest running efforts.
Further investment in individual level trait data for long-term monitoring will be critical for assessing finer scale shifts that likely occur more commonly in communities and potentially preceded bigger changes.

Third, while we use here the most common metrics for describing functional diversity there are a number of possible metrics for assessing a functional space, with new ones frequently proposed.
There may be a significant shift in a characteristic of the functional space that was poorly captured by our chosen metrics.
<!-- Further exploration of complementarity of traits... -->

*Implications for Ecological Process*

Our results highlight a gap in both empirical and theoretical work linking shifts in community composition to functional change.
While a large body of work assesses study-level insights into temporal functional responses (citations), a macroecological perspective illustrates the urgent need for future work exploring the factors important in determining functional change under different species change scenarios.
Due to the many previously discussed barriers to directly assessing functional shifts, a framework for predicting functional change from community characteristics would be a substantial step forward in understanding biodiversity change.

Much of the work linking species change (or loss) to ecological process either implicitly or explicitly assumes functional loss follows from species loss in natural communities.
For example ....
This assumption is clearly refuted by our results.

Our results have important implications for how we approach work linking species loss to ecosystem processes, which often implicitly assume concurrent functional loss.
There is no evidence of directional functional loss or even functional shifts accompany species loss in these natural communities.
<!-- clarify that this is about relationship between concurrent patterns in a single community --> Building on a rich body of work linking species diversity changes to changes in ecosystem function, resilience, and stability (citations), it will be essential to tease apart effects due to species change from those due to functional shifts if we want to infer impacts in natural communities.
--\>

<!-- call for theory development, punt back to community ecology - we don't know what this means! -->

Our results emphasize the importance of directly assessing functional change rather than assuming functional loss accompanies species loss.

-   getting functional data is costly, so we'd like to be able to say something about udnerlying functional composition without it

-   list of studies where they make functional inference without only species level data

Also illustrates the lack of theoretical development surrounding the relationship between functional and species change, and when you would expect different kinds of relationships.

-   also theory around the implications of functional shifts for ecological processes

**Notes**

@@

-   land use history and disturbance showing different relationships between species and functional diversity (Cadotte 2011)
